---
title: <center><font size="7"><b>Gráficos elegantes con ggplot2</b></font></center>
subtitle: <center><font size="4"><b>Programación y métodos estadísticos avanzados en R</b> </font></center>
author: <center><font size="5"><a href="http://marceloarayasalas.weebly.com/">Marcelo Araya-Salas, PhD</a></font></center>
date: <center>"`r Sys.Date()`"</center>
output:
  html_document:
    css: extra.css
    df_print: tibble
    highlight: pygments
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
fontsize: 12pt 
editor_options: 
  chunk_output_type: console
---


```{r, echo = FALSE, message=FALSE}

x <- c("kableExtra", "knitr", "ggplot2", "RColorBrewer", "knitr", "MASS", "viridis")

aa <- lapply(x, function(y) {
  if(!y %in% installed.packages()[,"Package"])  {if(!y %in% c("warbleR", "Rraven")) install.packages(y) else devtools::install_github(paste0("maRce10/", y))
}
try(require(y, character.only = T), silent = T)
  })

options("digits" = 3, "digits.secs" = 4, knitr.table.format = "html", results = 'asis') 


tibble <- function(x, ...) { 
  x <- kbl(x, digits=4, align= 'c', row.names = FALSE) 
   x <- kable_styling(x, position ="center", full_width = FALSE,  bootstrap_options = c("striped", "hover", "condensed", "responsive")) 
   asis_output(x)
}

registerS3method("knit_print", "data.frame", tibble)

# ggplot settings
geom_histogram <- function(...) ggplot2::geom_histogram(..., fill = viridis(10, alpha = 0.5)[8], show.legend = FALSE, bins = 20, color = "black")

geom_smooth <- function(...) ggplot2::geom_smooth(..., color = viridis(10,  alpha = 0.5)[8])

geom_boxplot <- function(...) ggplot2::geom_boxplot(..., fill = viridis(10, alpha = 0.5)[7])

theme_set(theme_classic(base_size = 20))

# options to customize chunk outputs
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors", # for code line numbers
  tidy.opts = list(width.cutoff = 65), 
  tidy = TRUE,
  message = FALSE
 )

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>


```{r functions and data, eval=T, echo=F}

hylaeformis_data <- read.csv("https://raw.githubusercontent.com/maRce10/OTS_TB_2022/master/data/hylaeformis_data.csv", stringsAsFactors = FALSE)

```

&nbsp; 

<div class="alert alert-warning">

## ggplot2

  - An R package specifically designed to produce graphics

  - Unlike other packages, ggplot2 has its own grammar

  - The grammar is based on "Grammar of Graphics" (Wilkinson 2005)

  - Independent modules that can be combined in many forms

  - This grammar provides high flexibility

</div>

&nbsp; 

## Grammar of graphics

The main idea is to start with a base layer of raw data and then add more layers of annotations and statistical summaries. The package allows us to produce graphics using the same structure of thought that we use when designing an analysis, reducing the distance of how we visualize a graphic in the head and the final product.

Learning the grammar will not only be crucial to produce a graph of interest, but also to think about other more complex graphs. The advantage of this grammar is the possibility  to create  new graphs composed of new combinations of elements.

&nbsp; 

## Graph components

All ggplot2 graphs contain the following components:

  -	**Data** - The R object with the information that needs to be plotted
  -	**layers** - The specific data that will be plotted (e.g. 'x' & 'y')
  -	**scale** - The range of the data to be included
  -	**coord.** - Coordinate system (not used very often)
  -	**facet** - determines how to break the data in subplots in a multipanel
  -	**theme** - controls the plot style 
&nbsp; 

This components are put together using "+".

The most common syntax includes the data within the "ggplot" call and a "geom_" layer.

&nbsp; 

First install/load the package:

```{r, eval=FALSE}

#install
install.packages("ggplot2") 

# load library
library(ggplot2) 

```
&nbsp; 

## Scatter plots

Let's use the "iris" data set to create scatter plots:

```{r}

ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Petal.Length)) + geom_point()

```
&nbsp; 

This plot is defined by 3 components:
  1. "data"- iris
  1. "aes" - Sepal.length vs Petal.length
  1. "layer" - Points (geom)

&nbsp; 

## Aesthetic attributes

We can also add other aesthetic attributes like color, shape and size. This attributes can be included within `aes()`:

```{r}

# color by species
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Petal.Length, color = Species)) + geom_point()

# color and shape by species
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Petal.Length, color = Species, shape = Species)) + geom_point()

```
&nbsp; 

Note that the aesthetic arguments can also be included in the "geom" layer:

```{r}

ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Petal.Length)) + geom_point(aes(color = Species, shape = Species))

```
&nbsp; 


We can also include a fixed value:

```{r}

ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Petal.Length)) + geom_point(color = "red2")

```
&nbsp; 

Some attributes work better with some data types:

  - Color and shape: categorical variables
  - Size: continuous variables

&nbsp; 


</br>

<div class="alert alert-info">

<font size="5">Exercise 1</font> 

Using the "hylaeformis" data set:

```{r,  eval=F, echo=T}

#read from website
hylaeformis_data <- read.csv("https://raw.githubusercontent.com/maRce10/Stats-Workshop-OTS_TB_2022-3/master/data/hylaeformis_data.csv", stringsAsFactors = FALSE)

# or  download manually and read from local file
hylaeformis_data <- read.csv("hylaeformis_data.csv", stringsAsFactors = FALSE)

head(hylaeformis_data, 20)

```

</br>

1.1 Create a scatter plot of "duration" vs "meanfreq" (mean frequency)

```{r, eval = FALSE, echo = FALSE}

ggplot(data = hylaeformis_data, mapping = aes(x = duration, y = meanfreq)) + geom_point()

```

</br>

1.2 Add a aesthetic attribute to show a different color for each locality

```{r, eval = FALSE, echo = FALSE}

ggplot(data = hylaeformis_data, mapping = aes(x = duration, y = meanfreq)) + geom_point(aes(color = Locality))

```

</br>

1.3 Add another aesthetic attribute to show "dfrange" (dominant frequency range) as the shape size

```{r, eval = FALSE, echo = FALSE}

ggplot(data = hylaeformis_data, mapping = aes(x = duration, y = meanfreq)) + geom_point(aes(color = Locality, size = dfrange))

```

</br>

</div>

&nbsp;

## Multipanel plots (Facetting)

  - Another way to visualize categorical variables
  - Allows to create multipanel plots for each level of the variable
  - 2 types: "grid" & "wrap"
  
```{r}

ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point() + 
  facet_wrap(~Species)

# or

ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point() + 
  facet_grid(~Species)

```
&nbsp; 


The scale can be fixed or free for the x and y axis, and the number of columns and rows can be modified:

```{r}
# free x
ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point() + 
  facet_wrap(~Species, scales = "free_x")
```


```{r, fig.height=8}
# free x and 3 rows
ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point() + 
  facet_wrap(~Species, scales = "free_y", nrow = 3)
```


```{r, fig.height=6}
# both free and 2 rows
ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point() + 
  facet_wrap(~Species, scales = "free", nrow = 2)

```
&nbsp; 

Note that we can also saved the basic component as an **R** object and add other components later in the code:

```{r, fig.height=8}

p <- ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point() 

p + facet_wrap(~Species, scales = "free_x", nrow = 3)

```


## Additional "geoms"

- `geom_smooth()` - adds best fit lines (including CI)
- `geom_boxplot()` 
- `geom_histogram()` & `geom_freqpoly()` - frequency distributions
- `geom_bar()` - frequency distribution of categorical variables
- `geom_path()` & `geom_line()` - add lines to scatter plots

&nbsp; 

### geom_smooth()

Best fit regression lines can be added with `geom_smooth()`:

```{r, fig.height=8}

# smoother and CI
ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  facet_wrap(~Species, scales = "free", nrow = 3)

# without CI
ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Species, scales = "free", nrow = 3)


```
&nbsp; 

<div class="alert alert-info">

<font size="5">Exercise 2</font> 

Using the "msleep" example data set:

</br>

2.1 Create a scatter plot of "bodywt"(body weight)  vs "brainwt" (brain weight)

```{r, eval = FALSE, echo = FALSE}

data(msleep)
ggplot(data = msleep, mapping = aes(x = bodywt, y = brainwt)) + geom_point()

```

</br>

2.2 Add "order" as a color aesthetic

```{r, eval = FALSE, echo = FALSE}

ggplot(data = msleep, mapping = aes(x = bodywt, y = brainwt, color = order)) + 
  geom_point()

```

</br>

2.3 Add a "facet" component to split plots by order using free scales

```{r, eval = FALSE, echo = FALSE}

ggplot(data = msleep, mapping = aes(x = bodywt, y = brainwt)) + 
  geom_point() +
  facet_wrap(~order, scales = "free")

```

</br>

2.4 Remove the orders with less than 4 species in the data set and make a plot similar to 2.3

```{r, eval = FALSE, echo = FALSE}

tab <- table(msleep$order)

ms2 <- msleep[msleep$order %in% names(tab)[tab >= 5], ]

ggplot(data = ms2, mapping = aes(x = bodywt, y = brainwt)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~order, scales = "free")

```

</br>

2.5 Add a smooth line to each plot in the panel

```{r, eval = FALSE, echo = FALSE}

ggplot(data = msleep[msleep$order %in% names(tab)[tab > 3], ], mapping = aes(x = bodywt, y = brainwt)) + 
  geom_point() +
  geom_smooth() +
  facet_wrap(~order, scales = "free")

```
</br>

</div>


### Boxplots

Again, it only takes a new "geom" component to create a boxplot:

```{r}

ggplot(iris, aes(Species, Petal.Length)) + geom_boxplot()

```

An interesting alternative are the violin plots:
```{r}

ggplot(iris, aes(Species, Petal.Length)) + geom_violin()

```
&nbsp; 

### Histograms

Same thing with histrograms and frequency plots:

```{r}

ggplot(iris, aes(Petal.Length)) + geom_histogram()

ggplot(iris, aes(Petal.Length)) + geom_freqpoly()

ggplot(iris, aes(Petal.Length)) + geom_histogram()  + geom_freqpoly()

```
&nbsp; 

We can control the width of the bars:
```{r}

ggplot(iris, aes(Petal.Length)) + 
  geom_histogram(binwidth = 1, fill = adjustcolor("red2", alpha.f = 0.3))

```
&nbsp; 

And compare the distribution of different groups within the same histogram:
```{r}

ggplot(iris, aes(Petal.Length, fill = Species)) + geom_histogram(binwidth = 0.4)

```
&nbsp; 

### Bar plots

Show the distribution of discrete (categorical) variables:

```{r}

tab <- table(msleep$order)

df <- as.data.frame(table(msleep$order[msleep$order %in% names(tab)[tab > 3]]))

ggplot(df, aes(Var1, Freq)) + geom_bar(stat = "identity")

```
&nbsp; 


```{r, eval=F, echo=F}

# Calcular el promedio del # de reclutas por estación
df2 <- aggregate(RECRUITS ~ SEASON, df, mean)
names(df2) <- c("season", "rec")  

# Gráfico de barras
p <- ggplot(df2, aes(season, rec)) +
  geom_bar(stat="identity")
p
   
# Cambiar la orientación
p + coord_flip()

### Cambiar el ancho y colores de las barras ###

# Cambiar el ancho de las barras
ggplot(df2, aes(season, rec)) +
  geom_bar(stat="identity", width = 0.5)

ggplot(df2, aes(season, rec)) +
  geom_bar(stat="identity", width = 0.2)

# Cambiar colores
ggplot(df2, aes(season, rec)) +
  geom_bar(stat="identity", color = "blue", fill="white")

# Remover el tema y agregar barras azules
p <- ggplot(df2, aes(season, rec)) +
  geom_bar(stat="identity", fill="steelblue") + theme_minimal()
p

# Escoger que barras mostrar
p + scale_x_discrete(limits = c("Spring", "Winter"))

### Agregar anotaciones ###

df2$rec <- round(df2$rec, 1)

# Fuera de las barras
ggplot(df2, aes(season, rec)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = rec), vjust = -0.3, size = 3.5) +
  theme_minimal()

# Dentro de las barras
ggplot(df2, aes(season, rec)) +
  geom_bar(stat = "identity", fill = "yellow") +
  geom_text(aes(label = rec), vjust = 1.6, size = 3.5) +
  theme_minimal()

### Cambiar colores de barras por grupo ###

# Cambiar el color de las líneas externas por grupo
p <- ggplot(df2, aes(season, rec, color = season)) +
  geom_bar(stat = "identity", fill = "white")
p

### Cambiar el color manualmente ###

# Usar "palette" predefinido de colores
p + scale_color_manual(values = c("red", "blue", "green", "yellow"))

# Usar "palette brewer"
p + scale_color_brewer(palette = "Dark2")
p + scale_color_brewer(palette = 1)
p + scale_color_brewer(palette = 2)

# Usar escala de grises
p + scale_color_grey() + theme_classic()

### Cambiar colores de relleno de las barras ###

# Color de relleno
p <- ggplot(df2, aes(season, rec, fill = season)) +
  geom_bar(stat = "identity") + theme_minimal()
p

### Cambiar el color de relleno manualmente ###

# Usar "palette" predefinido de colores
p + scale_fill_manual(values = c("red", "blue", "green", "yellow"))

# Usar "palette brewer"
p + scale_fill_brewer(palette = "Dark2")
p + scale_fill_brewer(palette = 1)
p + scale_fill_brewer(palette = 2)

# Usar escala de grises
p + scale_fill_grey() + theme_classic()

# Usar color negro del borde y relleno
ggplot(df2, aes(season, rec, fill = season)) +
  geom_bar(stat = "identity", color = "black") + 
  scale_fill_manual(values = c("red", "blue", "green", "yellow")) +
  theme_classic()

### Cambiar la posición de la leyenda ###

# Cambiar el color de las barras a tonos de azúl
p <- ggplot(df2, aes(season, rec, fill = season)) +
  geom_bar(stat = "identity") + theme_minimal()

p1 <- p + scale_fill_brewer(palette = "Blues")
p1 + theme(legend.position = "top")
p1 + theme(legend.position = "bottom")

# Remover la leyenda
p1 + theme(legend.position = "none") 

### Cambiar el órden de los items en una leyenda ###

# Cambiar órden de grupos
p1 + scale_x_discrete(limits = c("Summer", "Autumn", "Winter", "Spring"))

### Editar aún más los elementos de la leyenda ###

# Cambiar el título de la leyenda y la fuente
p1 <- ggplot(df2, aes(season, rec, fill = season)) +
  geom_bar(stat = "identity", color = "black")

titulo <- "Estación"

p2 <- p1 + scale_fill_manual(titulo, values = c("red", "blue", "green", "yellow")) +
  theme_classic()

p2 + theme(legend.title = element_text(colour = "blue", size = 10,
                                       face = "bold"))

p2 + theme(legend.text = element_text(colour = "blue", size = 10,
                                       face = "bold"))

# Cambiar el color del fondo de la leyenda
p2 + theme(legend.background = element_rect(fill = "lightblue",
                                            size = 0.5, linetype = "solid"))

p2 + theme(legend.background = element_rect(fill = "lightblue",
                                            size = 0.5, linetype = "solid"))

### Gráfico de barras con múltiples grupos ###
head(df)

df.n <- aggregate(RECRUITS ~ SEASON + DENSITY, df, mean)
names(df.n) <- c("season", "density", "rec")

# Gráfico 1
ggplot(df.n, aes(season, rec, fill = density)) +
  geom_bar(stat="identity")

# Gráfico 2
ggplot(df.n, aes(season, rec, fill = density)) +
  geom_bar(stat="identity", position=position_dodge())

# Cambiar el color manualmente
p <- ggplot(df.n, aes(season, rec, fill = density)) +
  geom_bar(stat="identity", color = "black", position=position_dodge()) +
  theme_minimal()

p + theme_classic()

# Colores manuales
p + scale_fill_manual(values = c('#999999','#E69F00')) + theme_classic()

# Palette de colores
p + scale_fill_brewer(palette = "Blues") + theme_classic()

### Gráficos de barras con error ###

# Calcular promedio +/- SD
head(df)

df3 <- aggregate(RECRUITS ~ SEASON + DENSITY, df, FUN = function(x) c(mean = mean(x), sd = sd(x)))
df4 <- cbind(data.frame(df3[,c(1:2)]), df3$RECRUITS[,1], df3$RECRUITS[,2])

names(df4) <- c("season", "density", "mean", "sd")
str(df4)

# Grafico de barras y desviación estándar
p <- ggplot(df4, aes(season, mean, fill = density)) + 
  geom_bar(stat="identity", position=position_dodge(), color = "black") +
  geom_errorbar(aes(ymin = mean, ymax = mean + sd), width = 0.2,
                position = position_dodge(0.9))
  
p + scale_fill_brewer(palette="Paired") + theme_minimal()
p + scale_fill_brewer(palette="Paired") + theme_classic()

### Personalizar aún más los gráficos ###

# Poner títulos de ejes
p + xlab("Estación") + 
  ylab("Número de reclutas") +
  theme_classic()

# Remover títulos de ejes
p + xlab(NULL) + 
  ylab(NULL) +
  theme_classic()

# Otros cambios
p2 <- p + labs(x = "Estación", y = "Número de reclutas") + 
  scale_fill_manual(values = c("black", "grey50")) +
  theme_classic()
p2
 
# Cambiar los límites X y Y
p2 + ylim(0,100)

# Cambiar límites con expand_limits
p2 + expand_limits(x = c(0, 6), y = c(0, 80))

```

## Customizing ggplots

Besides the basic functions (e.g. components) described above, ggplot has many other tools (both arguments and additional functions) to further customize plots. Pretty much every thing can be modified. Here we see some of the most common tools.

&nbsp; 

### Themes

ggplot2 comes with some default themes that can be easily applied to modified the look of our plots:

```{r}

p <- ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point() 

p + theme_classic()

p + theme_bw()

p + theme_minimal()
```
&nbsp; 

Most themes differ in the use of grids, border lines and axis labeling patterns.

&nbsp; 

### Axis customization

Axis limits can be modified as follows:

```{r}

ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point()  + 
  xlim(c(0, 10))

ggplot(iris, aes(Sepal.Length, Petal.Length, col = Species)) + 
  geom_point()  + 
  xlim(c(0, 10)) + 
  ylim(c(0, 9))

```
&nbsp; 

Axis can also be transformed:

```{r}

ggplot(iris, aes(Sepal.Length, Petal.Length, col = Species)) + 
  geom_point()  + 
scale_x_continuous(trans = "log") + 
  scale_y_continuous(trans = "log2")

```

&nbsp; 

or reversed:

```{r}

ggplot(iris, aes(Sepal.Length, Petal.Length, col = Species)) + 
  geom_point()  + 
scale_y_reverse() 

```



### Saving ggplots

ggplots can be exported as image files using the `ggsave` function:

```{r}

ggplot(data = msleep[msleep$order %in% names(tab)[tab > 5], ], mapping = aes(x = bodywt, y = brainwt)) + 
  geom_point() +
  facet_wrap(~order, scales = "free")

# Export
ggsave("plot.png", width = 5, height = 5)

```
&nbsp; 



The image file type will be identify by the extension in the file name

&nbsp; 


Additional axis customizing:

```{r}

# Log2 scaling of the y axis (with visually-equal spacing)
require(scales)

p + scale_y_continuous(trans = log2_trans())

# show exponents
p + scale_y_continuous(trans = log2_trans(),
    breaks = trans_breaks("log2", function(x) 2^x),
    labels = trans_format("log2", math_format(2^.x)))

# Percent
p + scale_y_continuous(labels = percent)

# dollar
p + scale_y_continuous(labels = dollar)

# scientific
p + scale_y_continuous(labels = scientific)

### Agregar "tick marks" ###

# Cargar librerías
library(MASS)

data(Animals)

# x and y axis are transformed and formatted
p2 <- ggplot(Animals, aes(x = body, y = brain)) + geom_point(size = 4) +
     scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x))) +
     scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x))) +
     theme_bw()

# log-log plot without log tick marks
p2

# Show log tick marks
p2 + annotation_logticks() 

# # Log ticks on left and right
p2 + annotation_logticks(sides = "lr")

# All sides
p2 + annotation_logticks(sides = "trbl")

```

&nbsp; 

### Other graphs 

Many other types of graphs can be generated. Here I show a single example of cool contour and "heatmap" graphs:

```{r}

head(faithful)

ggplot(faithfuld, aes(eruptions, waiting)) + geom_contour(aes(z = density, colour = ..level..))

ggplot(faithfuld, aes(eruptions, waiting)) + geom_raster(aes(fill = density))

```
&nbsp; 


## Other R graphing packages

-	**ggvis** (interactive ggplots)
-	**vcd** (Warnes 2015)
-	**plotrix** (Lemon et al. 2006)
-	 **gplots** (Warnes 2015)

Check the [CRAN Graphics Task View](http://cran.r-project.org/web/views/Graphics.html) for a more comprehensive list of graphical tools in **R**.

---

&nbsp; 

## References 

- Lemon J (2006) Plotrix: a package in the red light district of R. R-News
6(4):8–12
- Warnes GR, Bolker B, Bonebakker L, Gentleman R, Liaw WHA, Lumley
T, Maechler M, Magnusson A, Moeller S, Schwartz M, Venables B (2015) gplots: various R programming tools for plotting data. R package version 2.17.0. https://CRAN.R-project.org/package=gplots
- Wickham H (2010) A layered grammar of graphics. J Comput Graph Stat 19(1):3–28
- Wilkinson L (2005) The grammar of graphics. Statistics and computing, 2nd edn. Springer, New York

---

---

<font size="4">Session information</font>

```{r session info, echo=F}

sessionInfo()

```

